Under-ice acoustic navigation using real-time model-aided range estimation

### 2022-02-15 manuscript/main.tex
working through restructuring manuscript

\begin{abstract}
The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
For real-time navigation, this conversion assumes an isovelocity sound speed.
For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
This work demonstrates an embedded, ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
\llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
The vehicle was successfully deployed and recovered.
\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the localization performance is quantified between GPS-linked beacon-to-beacon connections where a single beacon ran the vehicle's navigation stack.
The real-time positioning error is roughly 11 meters at distances up to 3 km.
In post-processing, the dataset is extended to all possible beacon-to-beacon events.
The localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and locally observed sound speed profiles for a Minimal Bounce Criteria (MBC), as deployed in real-time, and an improved algorithm, the Nearest Bounce Criteria (NBC).
The best performing median re-positioning errors are roughly 10 and 2 meters, respectively.
\llabel{2.3} Further investigation suggests that the GPS ground truth data have some noise that is not reciprocated in the travel time data.
\end{abstract}

@@

\section{Introduction}
\label{sec:1}  
Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and samlpe the ocean, particularly for remote and/or dangerous environments.
However, navigation uncertainty is a major challenge in considering AUVs as reliable and standard tools for oceanographic research.
Compared to their land and air-based counterparts, underwater vehicles face a unique set of challenges brought on by the water column itself:  while land and air-based robots utilize the Global Navigation Satellite System (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GPS while underwater due to the rapid attenuation of electromagnetic waves \citep{preisig_acoustic_2007}.
Typically, an AUV can stall on the surface to receive a GPS fix, but this foolproof method of re-positioning is inaccessible in an ice-covered environment or other GPS-denied situations.

% Long baseline navigation
Though there many acoustic positioning methods, the long baseline (LBL) is closest in style

% A recent study tracking 22 floats under the ice in the Weddell Gyre near Antarctica showed that they were unable to surface for 8 months and position uncertainty grew to 116 $\pm$ 148 km in the lateral directions \citep{chamberlain_observing_2018}.

Much progress has been made in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.

\subsection{Timing, Positioning, and Navigation}

Underwater acoustic navigation (and GPS) is predicated on timing and positioning.
Timing is the ability to maintain accurate and precise time, sometimes synchronized across multiple nodes.
\llabel{1.11} The advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016} enabled ranging via one-way travel time (OWTT).

Positioning is the ability to accurately determine one's location and is generally done via trilateration.
Navigation assumes a degree of autonomy to move from a current position to a desired one.
The standard for real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
Short range (less than ten kilometers) navigation efforts achieve minimal navigation error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
For mesoscale (tens to hundreds of kilometers) navigation efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used, but error remains in the hundreds of meters.

Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}. 
The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.

\subsection{Underwater positioning and navigation in the Arctic}

Underwater navigation research is broadly motivated by GPS-like navigation in GPS-denied conditions.
The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.

Literature and historical reviews of underwater navigation approaches in under-ice conditions highlight the emerging technology learning curve for vehicle operations in the Arctic, the hostile conditions brought on by a semi-permanent moving ice cover, and the need for both absolute and relative navigation with little to no tolerance for error \citep{norgren_unmanned_2014,mcfarlane_autonomous_2015,barker_scientific_2020}.

One of the first successful AUV launches and recoveries involved equipping the vehicle with barbs which became entangled into a homing net underneath the ice \citep{bellingham_auv_1993}.
Since then, AUVs have been used for ocean bottom \citep{jakuba_long-baseline_2008,kunz_deep_2008}, under-ice \citep{kaminski_12_2010}, and hydrographic surveys \citep{kukulya_under-ice_2010,kukulya_development_2016,stevens_linking_2016,plueddemann_autonomous_2012}.
These efforts show a combination of short and long range positioning systems, where all acoustic navigation methods used a depth-averaged sound speed for linear scaling between recorded travel time and range.
In many cases robust quantitative navigation performance could not be evaluated due to small operational scale and/or moving ice floes.
But most importantly, all of these efforts experienced a nearly isovelocity or classical upward refracting Arctic acoustic environment.

The new Arctic\textemdash one with much less multi-year sea ice\textemdash is experiencing significant changes in the water column as well \citep{mackinnon_warm_2021}.
\llabel{2.4} The Beaufort Lens is a highly dynamic, highly stratified water mass, keeping a neutrally buoyant layer of warm Pacific water at 50 to 70 m in depth.
Surrounded by cooler waters, it distinguishes a local maximum in the SSP, creating an upper and lower duct\textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}. 
\llabel{2.5} Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \cite{kucukosmanoglu_observations_2021,badiey_azimuthal_2019,ballard_temporal_2020} indicate a high degree of spatio-temporal variability of the lower duct throughout the Chukchki and Beaufort Seas.

An experiment in 2014 exploited this lower duct to provide acoustic navigation and communication at mesoscale ranges in the Canada Basin.
\citet{freitag_long_2015} and \citet{webster_towards_2015} deployed multiple GPS-linked ice buoys with transducers that were capable of communications greater than 400 km with a range accuracy of 40 m when the sound speed was known.
The transducers were at 100 m depth, well-positioned in the lower duct spanning 50 to 200 m.
The glider navigation accuracy, however, was severely degraded as acoustic packets were sent when the gliders were outside the sound channel.
When gliders were deployed again in 2017, their navigation error was on the order of half a kilometer, and a glider re-navigation using an acoustic arrival method and the average sound-speed profile from the gliders reduced navigation error by a factor of 4-5, depending on the dive \citep{graupe_preliminary_2019}.
Navigation error is referenced to intermittent GPS fixes, as all of these glider deployments occurred in the Marginal Ice Zone with little to no ice at the time.

\subsection{Ice Exercise 2020}

The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more inaccurate due to increased multipath in the upper duct of the Beaufort Lens.

The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
The challenges anticipated for ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.

In this paper we present a working solution for underwater navigation that achieves GPS-like accuracy and precision.
We embed a stochastic prediction of a single group velocity to convert recorded travel time into range.
This prediction ingests an updatable snapshot of the SSP via acoustic message; over the course of ICEX20, this capability was used several times. 
Ultimately, integrating environmental realism enabled a successful emergency recovery of the AUV.
Due to a harddrive error, the AUV stalled underneath the ice surface and broadcasted its location.
A team was able to dig a hole through the 2 meter ice sheet and immediately see the vehicle.
Its successful recovery amidst total ice cover is shown in Fig. \ref{fig:vehicleRecovery}.
\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/Fig1.jpg}
	\caption{The successful emergency recovery of AUV Macrura, with the the auger used to drill through the ice sheet in the background. Photo courtesy of LT CDR D. Goodwin.}
	\label{fig:vehicleRecovery}
\end{figure}

The rest of this paper is organized as follows: Section~\ref{sec:methods} presents the physical
experimental setup and model-based data processing; 
section \ref{sec:realtime} quickly conveys real-time results and the limitations of the data collected; section~\ref{sec:postprocess} presents key results for group velocity predictions and range estimations; section~\ref{sec:discussion} closes with a discussion on importance, future work, and broader relevance.

% =========================================================================== %
% =========================================================================== %
\section{Methods}
\label{sec:methods}

The acoustic communication dataset accrued for ICEX20 relied on the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
Both simulation capabilities are integral to mission success.

\subsection{Operational paradigm on the ice}

The ICNN is comprised of four ice buoys, in a loose rectangle, roughly 2 km away from a central ice camp with a topside computer, as shown in Fig.~\ref{fig:icnnOverview}.

\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/Fig2.pdf}
	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
	\label{fig:icnnOverview}
\end{figure}

Each ice buoy is outfitted with a WHOI Micro-Modem \citep{singh_underwater_2006}, with a four-element receiver array and a single transmitter, and one-tenth of a millisecond resolution.
The receiving and transmitting elements were split between shallow and deeper depths (30 and 90 m, respectively) to provide better coverage due to the acoustic shadow zone in the Beaufort Lens.
The buoys do not encompass the full operating range of the vehicle but are positioned to minimize overlap in trilateration for spherical positioning \citep{deffenbaugh_relationship_1996}.

Vehicle missions in any underwater environment, but especially acoustically complex ones, must balance competing uses of the acoustic channel.
This network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
The ICNN operational paradigm is as follows:
\begin{enumerate}
\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011} coupled with a Virtual Ocean \citep{schneider_netsim_2018}
\item The topside computer calculates a new position by trilaterating the range estimates
\item The position differential, not the absolute position, is broadcast to the vehicle to update its navigation solution and be robust to latency and intermittency
\end{enumerate}

This paper specifically addresses the third step\textemdash the topside computer's conversion of travel time into range, which underpins the trilateration in the ICNN, and is an often overlooked element of any underwater acoustic navigation system.
The analysis accordingly focuses on validating acoustic range estimates from communication events between GPS-tracked beacons, not the AUV itself.
This allows the results to be decoupled from all the other moving parts in the ICNN.
This ``modem experiment'' was conducted while the vehicle was still being prepped to go under the ice.

\subsection{Modem experiment design}

Because the navigation solution on the vehicle during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
The ice buoy modems were run as ``virtual vehicles'' at a given depth and would transmit to all other ice buoys.
The internal sound speed estimate was then toggled using a handful of weights applied to a basis function representation of the sound speed.
This trial was configured for as many ice buoy modems as possible with the time allotted, with each experiment running for roughly an hour between the two sound speed modalities.
There are three important engineering choices in the ICNN network to contextualize this dataset:
\begin{enumerate}
\item Each modem buoy only has one transmit depth layer (30 or 90 m)
\item \llabel{1.12} The transmit and receive layers are independent
\item At any given time, the receive depth layers are the same across all four ice buoys
\end{enumerate}

The design of the ICNN enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
This adaptivity was not used during the modem experiment but was instead toggled manually.
Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were active during the experiment recorded events as tracking buoys, not as vehicles.

\llabel{1.13} A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.
Furthermore, because one of the buoys was configured to act as a virtual vehicle at any given time during the modem experiment, the tracking range was supplemented with an additional modem deployed from the camp site and configured to act as a virtual buoy; the transducer attached to the camp site modem was lowered to a depth of 20 m.

\subsubsection{Physical layer}
The physical layer describes the acoustic link between beacons in the water column.
It covers all source depths (20, 30, and 90 m) and both buoy receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.51\textwidth]{figs/Fig3a.pdf} \hfill
  \includegraphics[width=.48\textwidth]{figs/Fig3b.pdf} \\
  \vspace{1em}
  \includegraphics[width=.48\textwidth]{figs/Fig3c.pdf} \hfill
  \includegraphics[width=.48\textwidth]{figs/Fig3d.pdf} \\
  \caption{An overview of the modem experiment by source and receiver depth and position. Subfigure (a) shows a chart of the successful events between source and receiver depth. The black column on the left, $tx$, shows the source depth, $z_s$. The column on the right, $rx$, shows the receivers with the amount of good contacts. The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m. Subfigures (b-d) correspond to transmitter depths of 20, 30, and 90 m, respectively. Topside, demarcated by the purple triangle, is situated just north of the continental shelf.}
  \label{fig:overview}
  \end{figure}

\clearpage

\subsubsection{Virtual layer}
The virtual layer describes the BELLHOP simulations for each communication event in the physical layer.
The SSP used for the raytracing is driven by a handful of weights applied to a basis function representation
of the SSP variability to reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
The SSP is the only human configurable input to the group velocity prediction.

\llabel{1.14} During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a local CTD cast, i.e. an optimal set of weights chosen by a human decision maker \llabel{2.7} via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
\llabel{1.8} For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B. Howard and LT CDR D. Goodwin).
This order corresponds to increasing ducted conditions of the Beaufort Lens.

Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
\llabel{1.16} It is important to note that BELLHOP produces an \llabel{2.8}abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.

\begin{figure}[h!]
  \centering
  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m. The beacons are highlighted in color/marker coding in Fig. \ref{fig:overview}. The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
  \label{fig:raytrace-zs30}
  \end{figure}

\FloatBarrier
\subsection{Group velocity prediction methods}

The variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm, especially in an acoustically complex propagation environment.
The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
\llabel{2.9} For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.

\subsubsection{Minimal bounce criteria (MBC)}
Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e. the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.

\llabel{2.10} In the ICEX20 configuration, the acoustic tracking is running on the topside computer, which controls the integrated communication and navigation range.
Here we assume that the group velocities $u_{i,j}$ are smoothly varying over the course of a vehicle mission, i.e., with respect to range, mission time, and the frequency of updates relative to vehicle motion. 
The group velocity is continuously tracked using predictions from the \textit{Virtual Ocean} infrastructure.

\llabel{1.15} When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program. The latter returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.

\llabel{2.11} The initial call to BELLHOP is over a sparse local grid, from which the plane wave approximation is used to interpolate the ray to the range and depth posited by the onboard tracking solution.
For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
\begin{equation}
u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
\end{equation}
and the associated weighted standard deviation,
\begin{equation}
\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
\end{equation}
If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.

Finally, the pseudorange is calculated simply as
\begin{equation}
r_{i,j} = u_{i,j} \Delta t_{i,j} 
\end{equation}

This stochastic method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than general post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
The method assumes direct path arrival, where multipath over the shallow receiver depth is trivial for the group velocity calculation; thus we call it the minimal bounce criteria (MBC).
The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees.
However, this high density of launch angles creates a greedy estimator of direct path eigenrays, and this simulation pipeline almost always guarantees a direct path found.
The MBC improves upon a deterministic estimate in that it considers the simplest ray path, travel time, and amplitude to predict a realistic group velocity.

\subsubsection{Nearest Bounce Criteria}

\llabel{2.12} As shown in the eigenray traces of Fig. \ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
This metric was evaluated in post-processing.

Given a running estimate for the horizontal group velocity $u_{i,j}$ between nodes $i$ and $j$, the navigation system has an extrapolated value for range, $\hat{r}$, in the virtual layer, and a recorded travel time, $\Delta t_{i,j}$, in the physical layer.
Instead of using only the $N_0$ rays with neither surface nor bottom bounces to estimate group velocity and subsequently moving to incremental number of bounces only when no valid direct path solutions exist, we solve for the power weighted average of the ray travel time for the $N_k$ rays with $k$ bounces,
\begin{equation}
t_k = \frac{\sum_{n=1}^{N_{k}} dt_{n}a_{n}^{2}}{\sum_{n=1}^{N_{k}} a_{n}^{2}} ~, 
\end{equation}
find the nearest matching power weighted average to recorded travel time,
\begin{equation}
t_{i,j,k} = \min_{k=0,1,2,...} \left| t_k - \Delta t_{i,j} \right|
\end{equation}
predict a group velocity,
\begin{equation}
u_{i,j} = \dfrac{\hat{r}}{t_{i,j,k}}
\end{equation}
and estimate the range as was done previously.
\begin{equation}
r_{i,j} = u_{i,j}\Delta t_{i,j}
\end{equation}

This method selects a different group velocity based on the multipath arrival structure, as the detected arrival is not always the first arrival or the direct path and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations. 
\llabel{1.17} In addition, for a more consistent comparison, the local sparse grid is standardized as 11$\times$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:raytrace-zs30}.

\FloatBarrier
\section{\label{sec:realtime} Real-time pseudorange results}

The modem experiment generated 811 beacon to beacon communication events with their own real-time MBC group velocity predictions.
\llabel{1.18} As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
This section examines the absolute range error between the pseudorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.

\begin{table}[h!]
\renewcommand{\arraystretch}{1.3}
\centering
\begin{tabular}{r|c|c|c|c|c}
SSP & Count & ~Mean [m] & ~Median [m] & ~STD [m] & ~Max [m] \\ \hline
Baseline & 243 & 11.38 & 11.96 & 4.23 & 23.95 \\ 
Chosen weights & 568 & 11.36 & 11.40 & 8.12 & 26.0 \\
\toprule
\end{tabular}
\caption[Range estimation error for \textit{in situ} events]{The range estimation error metrics for \textit{in situ} events. The error here is presented as the absolute, not directional, error. Almost all directional errors, however, are positive.}
\label{tab:rangeErrorInSitu}
\end{table}

Table \ref{tab:rangeErrorInSitu} breaks down the number of events, mean, median, standard deviation, and maximum absolute range error for the \textit{in situ} calculations.
The mean and median for both sound speeds are similar on the order of centimeters; the chosen weights show a maximum 2 meters greater than that of baseline, but it is from a communication event that has no comparison in the baseline data set.
Analysis shows that where there is overlap in range between sound speed conditions used for the real-time MBC, the range error difference is no more than a few meters.
\llabel{2.14} This is likely driven by the prominence of the duct.

The overarching statistical results show that both sound speed environments, given the smaller operational scales and shallow depths, are accurate enough to approximate the group velocity structure given the local grid.
Indeed, the error statistics indicated here are similar to those reported for the ICNN \citep{randeni_high-resolution_2021} and accurate enough to have facilitated a successful vehicle recovery.

\FloatBarrier
\section{\label{sec:postprocess} Post-processing pseudorange results}

There are 1242 total beacon to beacon events in the modem experiment that would have been accepted by the ICNN for real-time ranging and navigation had the ranging behavior been running for all modems in the network, and not only for tracking the vehicle.
A post-processing analysis that replicated the real-time pipeline was run to overcome the unequal distribution of communication events with respect to depth, range, and sound speed status.

It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
The present analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.

\FloatBarrier
\subsection{Group velocity predictions}

>>>>

\begin{abstract}
The long baseline (LBL) underwater navigation paradigm relies on the conversion of recorded travel time to range to trilaterate for position.
For real-time operations, this conversion assumes an isovelocity sound speed.
For re-navigation in post-processing, computationally and/or labor intensive acoustic modeling may be employed to reduce uncertainty driven by multipath arrivals.
This work demonstrates a real-time ray-based prediction method of the effective horizontal group speed to minimize vehicle position error.
\llabel{2.1} This method was implemented for a small scale AUV-LBL system in March 2020, in the Beaufort Sea, in total under-ice conditions and a double ducted acoustic propagation environment.
The vehicle was successfully deployed and recovered.
\llabel{2.2} Given the lack of GPS data throughout the vehicle's mission, however, the localization performance between GPS-linked beacon-to-beacon connections serves as a proxy for the effectiveness of this approach.
The real-time positioning error between beacons was roughly 11 meters at distances up to 3 km, matching the magnitude of corrections recorded by the AUV during field operations.
In post-processing, the localization performance is reevaluated with an operationally equivalent navigation stack using modeled, historical, and locally observed sound speed profiles for a Minimal Bounce Criteria (MBC), as deployed in real-time, and an improved ray filtering algorithm, the Nearest Bounce Criteria (NBC).
The best performing median re-positioning errors by bounce criterias are roughly 10 and 2 meters, respectively.
\llabel{2.3} Further investigation suggests that the GNSS ground truth data have some noise that is not reciprocated in the travel time data.
Re-trilateration suggests that this approach effectively extends the single meter accuracy of the deployed GNSS pucks into the water column.
\end{abstract}

@@

\section{Introduction}
\label{sec:1}  
Autonomous underwater vehicles (AUVs) are increasingly capable platforms to explore and sample the ocean, particularly for remote and/or dangerous regions.
However, navigation uncertainty is a major challenge in considering AUVs as standard tools for oceanographic research.
While land and air-based robots utilize information from Global Navigation Satellite Systems (GNSS) to achieve stunning location accuracy and precision throughout the duration of their missions, AUVs cannot access GNSS while underwater due to the rapid attenuation of electromagnetic waves.
Therefore, underwater vehicles have relied on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and acoustic baseline positioning systems for navigation \citep{paull_auv_2014}.
Limiting navigation error and drift requires an AUV to periodically stall on the surface and obtain a GNSS fix to reset its position error.
This foolproof method of self-positioning is undesirable for stealth, adverse weather conditions, and mission efficiency, and inaccessible in a GPS-denied situation like an ice-covered environment.

% introduce LBL and precise statement of work
Of acoustic baseline navigation systems, long baseline (LBL) is the most GPS-like in style and scale, and most appropriate for mitigating drift without overburdening computation or payload size on the vehicle \citep{paull_auv_2014,van_uffelen_global_2021}.
The state-of-the-art for LBL outsources depth to a pressure sensor and solves the two-dimensional localization problem by assuming an isovelocity, linear scaling between one way travel time (OWTT) and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
This assumption is valid for short scale operations but oversimplifies propagation for larger and/or complex acoustic environments.
This paper demonstrates an embedded model-aided data processing approach to convert recorded OWTTs into pseudorange estimates.
This approach was necessitated by total under-ice conditions and a double ducted acoustic environment, in the Beaufort Sea, in March 2020, for an AUV deployment during the Ice Exercise 2020 (ICEX20).
We quantify the success of the proposed acoustic ranging method using GPS-linked beacon-to-beacon communication events, as the AUV-LBL positioning has no ground truth GNSS to compare to.

%% rest of the background, JASA style

% PNT
For clarity, we introduce definitions for timing, positioning, and navigation.

% More about LBL navigation
While RAFOS floats have championed one way ranging for re-positioning \citep{rossby_rafos_1986,duda_evaluation_2006}, the ability to do so for navigation was facilitated by \llabel{1.11} the advent of the WHOI micro-modem \citep{singh_underwater_2006} and synchronized chip scale atomic clocks \citep{gardner_second_2016}.
Several short range (less than ten kilometers) navigation efforts have achieved minimal error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
However, these efforts were all conducted with X sound speed conditions.
The approach in this paper, and other related works, was specifically designed to... the recently observed double ducted sound speed profile...
For convenience, we call this phenomena the Beaufort Lens, with the lens referring to ...

There is no comparable work in the Arctic for a short scale AUV deployment in the Beaufort Lens. 
Both seminal and more recent AUV deployments witnessed the classical upward refracting sound speed profile that is amenable to an isovelocity simplification.
Of note are large scale glider deployments; these, however, operated at X depths, X spatial scales, and did not need precise navigation because they experienced little to no sea ice.
At larger scales on the order of hundreds of kilometers, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed was used, but error remained in the hundreds of meters.
The sound speed value was chosen by...

The navigation approach proposed includes a non-deterministic relationship between travel time and range previously limited to large scale re-navigation and re-positioning efforts.
Such post-processing efforts can utilize additional information or computation that is inaccessible for onboard computation.

\begin{figure}[h!]
	\centering
	\includegraphics[width=\columnwidth]{figs/auv-track-update.png}
	\caption{The under-ice mission track for AUV Macrura, including the position updates as it stalled underneath the ice overnight. A marker was placed on the ice at the vehicle's estimated self-location. It was recovered after a three day storm within a meter of the marker.}
	\label{fig:vehicleRecovery}
\end{figure}
% include story later on and cite this figure?

\clearpage
\section{ICEX20 conditions and experiment design}

\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/ssp-gvel-large.pdf}
	\caption{Anticipated sound speed conditions: finish this using stuff from ICEX16, HYCOM, ITP}
	\label{fig:sspExpectation}
\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{figs/Fig2.pdf}
	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
	\label{fig:icnnOverview}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{figs/modem-chart.pdf}
  \caption{An overview of the modem experiment by source and receiver depth and position. The black column on the left, $tx$, shows the source depth, $z_s$. The column on the right, $rx$, shows the receivers with the amount of good contacts. The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m.}
  \label{fig:overview}
  \end{figure}

\clearpage
\section{\label{sec:realtime} Real-time pseudorange analysis}

\subsection{Minimal bounce criteria (MBC)}

\subsection{Pseudorange error metrics}

\subsection{Discussion}

\clearpage
\section{\label{sec:post} Post-processed pseudorange analysis}

\subsection{Inherent overestimation from the minimal bounce criteria}
\begin{figure}[h!]
  \centering
  \includegraphics[width=\reprintcolumnwidth]{figs/Fig4.pdf}
  \caption{Eigenrays for beacon to beacon events for each sound speed with a nominal source depth of 30 m. The beacons are highlighted in color/marker coding in Fig. \ref{fig:overview}. The eigenrays are curated from BELLHOP by travel time proximity and are traced in the representative receiver colors over a total ray fan in gray.}
  \label{fig:raytrace-zs30}
  \end{figure}

\subsection{Nearest bounce criteria (NBC)}

\subsection{Effective sound speed predictions}
